---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "85%"
)
```
# pedbuildr

The goal of pedbuildr is to reconstruct small/medium-sized pedigrees from genotype data. The most important functions of the package are

* `buildPeds()` : generates all pedigrees containing a given set of members
* `reconstruct()` (TO APPEAR): finds the most likely pedigree given the available genotype data

## Installation
The development version of pedbuildr is available from GitHub:
``` r
remotes::install_github("magnusdv/pedbuildr")
```
## Example: Building pedigree lists
```{r}
library(pedbuildr)
```

Suppose we are given 3 male individuals, labeled `1`, `2` and `3`. Using the `buildPeds()` function we create a list of all pedigrees containing these individuals:
```{r}
plist = buildPeds(ids = 1:3, sex = c(1, 1, 1))
length(plist)
```

Quite a large number! Let's see what the first 6 pedigrees look like. The function `plotPeds()` is a thin wrapper around `pedtools::plot.ped()`, and highlights the original individuals in each pedigree.

```{r fig.height=3, fig.width = 6, fig.align = "center"}
plotPeds(plist[1:6])
```

---

To reduce the number of pedigrees, we can put a restriction on the family structure. Suppose (e.g. after pairwise relatedness estimation) is known that individuals 1 and 2 form a parent-offspring pair, and likewise 1 and 3. In both cases, however, it is unknown which one is the parent. 

We convey this new piece of information by using the `knownPO` parameter:
```{r}
plist2 = buildPeds(1:3, sex = c(1, 1, 1), knownPO = list(c(1,2), c(1,3)))
length(plist2)
```

Here are the 15 pedigrees:
```{r fig.height = 6, fig.width = 8, fig.align = "center"}
plotPeds(plist2)
```
Note: The dotted curves indicate identical individuals. Hence there is a lot of inbreeding/incest in the pedigrees shown above.

## Which pedigrees are included in the `buildPeds()` algorithm?
TODO
