---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center"
)
```
# pedbuildr

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/pedbuildr)](https://CRAN.R-project.org/package=pedbuildr)
[![](https://cranlogs.r-pkg.org/badges/grand-total/pedbuildr?color=yellow)](https://cran.r-project.org/package=pedbuildr)
[![](https://cranlogs.r-pkg.org/badges/last-month/pedbuildr?color=yellow)](https://cran.r-project.org/package=pedbuildr)
<!-- badges: end -->

The goal of **pedbuildr** is to reconstruct pedigrees from genotype data. This is done by optimising the likelihood over all possible pedigrees subject to given restrictions. As part of the **ped suite** ecosystem of R packages for pedigree analysis, it uses on [pedtools](https://CRAN.R-project.org/package=pedtools) for handling pedigrees, and imports [pedprobr](https://CRAN.R-project.org/package=pedprobr) for calculating pedigree likelihoods.

See also the `ibdEstimate()` function of [forrel](https://CRAN.R-project.org/package=forrel), which does pairwise relatedness estimation.

## Installation
The **pedbuildr** package can be installed from CRAN as follows:
```{r, eval = FALSE}
install.packages("pedbuildr")
```

Alternatively, the latest development version is available from GitHub.
```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("magnusdv/pedbuildr")
```

## A reconstruction example
The built-in dataset `trioData` contains simulated genotypes for three individuals at 100 SNP markers. As a simple demonstration we will try to reconstruct the pedigree connecting these individuals. 

To get started, load **pedbuildr**.
```{r}
library(pedbuildr)
```

The `trioData` is a data frame in so-called *ped format*. Here are the first few columns:

```{r}
trioData[, 1:10]
```

We convert the data into a pedigree object, using the `as.ped()` function from **pedtools**.
```{r}
x = as.ped(trioData, locusAttributes = "snp12")
summary(x)
```
The `locusAttributes` argument tells R that all the markers are diallelic with alleles 1 and 2. By default, the alleles have equal frequencies.

To reconstruct the pedigree, simply run `reconstruct()`:
```{r}
res = reconstruct(x)
```

A tailor-made `plot` function makes it easy to visualise the most likely pedigrees:
```{r reconstruct-trio, fig.height = 4, fig.width = 8}
plot(res, top = 6)
```

## Further options
*A priori* there are infinitely many possible pedigrees connecting a set of individuals. (For example, two individuals may be *k*'th cousins for any *k* = 1,2,... .) In order to obtain a manageable search space, `reconstruct()` offers a range of restriction parameters:

* `extra`: The number of extra individuals allowed to connect the original individuals. (See further explanations below.)
* `age`: A numerical age vector, or a character vector describing age inequalities. For example, `age = c("A > B,C", "B > D")` excludes B and C as ancestors of A, and D as an ancestor of B (but no assumption is made for C vs. D).
* `inferPO`: If TRUE, an initial stage of pairwise IBD estimation is done, in order to infer certain parent-child pairs, as well as certain non-parent-child pairs. 
* `knownPO`: Known parent–offspring pairs.
* `notPO`: Pairs known not to be parent–offspring.
* `allKnown`: If TRUE, then `knownPO` is the complete list of parent–offspring pairs.
* `noChildren`: Individuals known to have no children.
* `maxInbreeding`: The highest permitted inbreeding coefficient in the pedigree. By default this is set to 1/16, which is appropriate for most projects involving human pedigrees.
* `linearInb`: The allowed level of inbreeding between linear descendants. For example, `linearInb = 1` allows mating between parent–child, but not grandparent–grandchild. The default value `FALSE` disallows all inbreeding of this type.
* `connected`: If TRUE (default), only connected pedigrees are considered.
* `sexSymmetry`: If TRUE (default) pedigrees are considered equal if they differ only in the
sexes of the added parents, e.g., paternal versus maternal half-siblings.

Let us re-run the reconstruction of `trioData` adding a few of these restrictions. We allow 3 extra individuals and indicate that individual 1 is older than the others. Furthermore, we ask the program to infer parent-child relationships automatically, and disallow linear inbreeding.
```{r}
res2 = reconstruct(x, extra = 3, age = "1 > 2,3", inferPO = TRUE)
```
The most likely results this time are shown below:
```{r reconstruct-trio-2, fig.height = 4.3, fig.width = 8}
plot(res2, top = 6)
```

We see that the same pedigree "wins", but some esoteric alternatives have appeared among the runners-up.

## More about `extra`
Arguably the most important parameter to `reconstruct()` is `extra`, which controls the size of the pedigrees to consider. It can be either a nonnegative integer, or the word "parents". If an integer, it sets the maximum number of extra members used to connect the original individuals. (The final pedigrees may contain further extras still, since missing parents are added at the end.)

If `extra = "parents"`, a special algorithm is invoked. First all directed acyclic graphs between the original individuals are generated, and then parents are added in all possible ways. This is (currently) the default behaviour, since it avoids setting an *ad hoc* number of "extras". However, it only works well in relatively small cases.
